---
title: 'Reproducible Research Course Assignment: Prj. 1'
author: "Menelik G. Wondimu"
date: "September 8, 2018"
output: html_document
---

#### Load all the required libraries 
```{r setup, echo=TRUE, include= FALSE}
library(ggplot2)  
library(mice)
library(randomForest)
library(lattice)
library(plyr)
```

#### Download the project data to the set directory and unzip the file

```{r download, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, results='hide'}
URL<-'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip'
download.file(URL, destfile = 'C:\\Users\\asten\\Documents\\Mini\\Data_science_training\\Reproducible_Research\\week2\\CourseProject1\\Activity_Monitoring_Data.zip', method='auto')
unzip('Activity_Monitoring_Data.zip')
```

#### Load the Project dataset

```{r Load_the_data, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, results='hide'}
Activity_Monitoring_Data <- read.csv("activity.csv")
```

#### **1. Calculate the total number of steps**

```{r Total_Number_of_Steps, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, results='hide'}
Total_No_of_Steps<-aggregate(Activity_Monitoring_Data['steps'], by=Activity_Monitoring_Data['date'], sum)
```

#### **1.1. Make a histogram of the total number of steps taken each day**

```{r plot_histogram, echo= TRUE,include= TRUE}
hist(Total_No_of_Steps$steps,freq = TRUE,
     xlim=c(0,25000), ylim=c(0,30),
     xlab = "Total Number of Steps Taken Per Day", 
     ylab = "Frequency", main = "Activity Histogram", 
     col = "darkblue", border = "red")
```

####**1.2. Calculate and report the mean and median of the total number of steps taken per day.**

```{r Total_Noof_Steps,echo=TRUE,include= TRUE}
summary(Total_No_of_Steps$steps)
```

####As can be seen from the summary table above the mean is 10766 and the median is 10765

#### **2. Make a time series plot (i.e. type = "l"') of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)**

```{r Time_series_data, echo= TRUE, results='hide'}
Avg_interval_steps<-with(Activity_Monitoring_Data,tapply(steps,interval,mean,na.rm=TRUE))
Intervals<-unique(Activity_Monitoring_Data$interval)
New_data<-data.frame(cbind(Avg_interval_steps,Intervals))
```

#### The time series plot of the 5-minute interval steps

```{r Time_series_plot,echo= TRUE,include= TRUE}
plot(New_data$Intervals,New_data$Avg_interval_steps,type = "l",xlab = "Intervals",
     ylab = "Average Steps",main = "Average Steps per Interval")
```

#### **2.1. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?**

```{r Maximum_Number_of Steps, echo= TRUE, results='hide'}
Max_no_steps<- max(New_data$Avg_interval_steps)
```

```{r Intervl containing maximum number of steps, echo= TRUE, include= TRUE}
subset(New_data, Avg_interval_steps == max(Avg_interval_steps))
```


#### **3. Calculate and report the total number of missing values in the dataset.**

```{r Total number of missing values, echo= TRUE, include= TRUE}
sum(is.na(Activity_Monitoring_Data)) 
```

#####Hence, from the above result the total number of missing values is 2304

#### **3.1 Create a new dataset that is equal to the original dataset but with the missing data filled in**

#### My strategy for inputting missing data is perform mice imputation, based on random forests. MICE operation is a simplified approach to impute missing data with MICE package in R.

```{r Perform mice imputation based on random forests, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, results='hide'}
New_Dataset <- mice(Activity_Monitoring_Data[, !names(Activity_Monitoring_Data) %in% "medv"], method="rf")
```

```{r Generate the completed data,echo= TRUE, results='hide'}
New_Act_Dataset <- complete(New_Dataset)  
```

```{r New dataset with the missing data filled in,echo= TRUE, results='hide'}
New_Total_Steps<-aggregate(New_Act_Dataset['steps'], by=New_Act_Dataset['date'], sum)
```

#### **3.2. Make a histogram of the total number of steps taken each day**

```{r plot histogram of the total number of steps,echo= TRUE, include= TRUE}
hist(New_Total_Steps$steps,freq = TRUE,
     xlim=c(0,25000), ylim=c(0,30),
     xlab = "Total Number of Steps Taken Per Day", 
     ylab = "Frequency", main = "Activity Histogram", 
     col = "green", border = "red")
```

#### **3.3. Calculate and report the mean and median total number of steps taken per day.**

```{r Mean and Median total number of steps,echo= TRUE, include= TRUE}
summary(New_Total_Steps$steps)
```

#### From the above summary, the mean and median is found to be 9530 and 10395 respectively.   

#### **4. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.**

```{r Create a new factor variable,echo= TRUE, results='hide'}
New_Act_Dataset$date <- as.Date(New_Act_Dataset$date)
```

```{r create a vector of weekdays,echo= TRUE, results='hide'}
weekdays1 <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
```

#### Use `%in%` and `weekdays` to create a logical vector and onvert to `factor` and specify the `labels/levels`

```{r create logical vector, echo= TRUE, results= "hide"}
New_Act_Dataset$wDay <- factor((weekdays(New_Act_Dataset$date) %in% weekdays1),levels=c(FALSE, TRUE),labels=c('weekend', 'weekday')) 
```

#### **4.1. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis)**

#### Calculate verage number of steps taken, averaged across all weekday days or weekend days

```{r AveregeSteps,echo= TRUE, results= "hide"}
AverageSteps <- ddply(New_Act_Dataset, .(interval, wDay), summarize, Avg = mean(steps))
```

#### The time series plot of weekday and weekend days is as shown below

```{r Timeseries plot,echo= TRUE, include=TRUE}
xyplot(Avg~interval|wDay, data = AverageSteps, type = "l", layout = c(1,2), main="Average Steps per Interval", ylab="Average number of Steps", xlab="Interval")
```


